---
title: "Polynomial Regression"
author: "Math 430, Winter 2017"
output:
  ioslides_presentation
---

## Modeling salary

```{r, include=FALSE}
library(ggplot2)
library(car)

profsalary <- read.table("http://www.stat.tamu.edu/~sheather/book/docs/datasets/profsalary.txt", header = TRUE)
```

What if the relationship between salary and years of experience wasn't linear?
    
```{r, message=FALSE, echo=FALSE}
ggplot(data = profsalary, aes(x = Experience, y = Salary)) +
  geom_point() + 
  geom_smooth()
```


## Polynomial regression



```{r, message=FALSE, echo=FALSE, fig.height=4, fig.width=5}
ggplot(data = profsalary, aes(x = Experience, y = Salary)) +
  geom_point() + 
  geom_smooth()
```

A quadratic model seems sensible.
$$Y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + e_i$$

## Polynomial regression in R {.smaller}

```{r}
quad.mod <- lm(Salary ~ Experience + I(Experience^2), data = profsalary)
summary(quad.mod)
```

## Interpreting coefficients

## Interpreting coefficients

## Polynomial regression {.smaller}

```{r fig.height=4, fig.width5}
ggplot(data = profsalary, aes(x = Experience, y = Salary)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2))
```

## Model checking

```{r, eval=FALSE}
residualPlots(quad.mod, type = "rstudent", pch = 16)
```


```{r, fig.width = 8.5, echo=FALSE, fig.height = 3}
residualPlots(quad.mod, type = "rstudent", pch = 16, layout = c(1, 3), tests = FALSE)
```

## Model checking

```{r, fig.width=5, fig.height = 5}
qqPlot(quad.mod, pch = 16, dist = "norm", reps = 5000)
```


## Beyond second-order models

Polynomials for one predictor

$$Y = \beta_0 + \beta_1 x + \beta_2 x^2 + \cdots + \beta_p x^p + \varepsilon_i$$
