---
title: "SLR Diagnostics in R"
author: "Math 430, Winter 2017"
output: slidy_presentation
---

# Set up for diagnostics

Load the R packages

```{r message=FALSE}
library(ggplot2) # for graphics
library(mosaic)  # alternative graphics
library(broom)   # for the augment() function
library(dplyr)   # for the arrange() function
```

Load the data set

```{r}
voltage <- read.csv("https://github.com/math430-lu/data/raw/master/voltage.csv")
head(voltage)
```


# Industrial experiment

- Batches of electrical insulating fluid were subjected to constant voltages until the insulating property of the fluids broke down. 

- 7 different voltage levels were studied and the measured responses were the times until breakdown.

```{r}
head(voltage)
```

# Fit the SLR model

```{r}
mod <- lm(Time ~ Voltage, data = voltage)
summary(mod)
```

# Plot the SLR model

```{r}
ggplot(data = voltage, aes(x = Voltage, y = Time)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Voltage applied (kV)", y = "Time until breakdown (min)")
```


# Getting ready for diagnostics

```{r}
diag_df <- augment(mod)
head(diag_df)
```

# Quick residual plots with `mplot`

```{r}
mplot(mod)
```


# Diagnostic plots the ggplot2 way

### Drawing a residual plot

```{r}
ggplot(data = diag_df, aes(x = .fitted, y = .std.resid)) +
  geom_hline(yintercept = 0, color = "blue") +
  geom_point() +
  labs(x = "Fitted value", y = "Standardized residuals")
```


# Diagnostic plots the ggplot2 way

### Location-scale plot

```{r}
ggplot(data = diag_df, aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(x = "Fitted value", y = expression(sqrt(abs("Standardized residual"))))
```



# Diagnostic plots the ggplot2 way

### Q-Q plot of the residuals

```{r}
ggplot(data = diag_df) +
  stat_qq(aes(sample = .std.resid))
```




# Plotting leverage

```{r}
# Flag leverage > 4/n
diag_df$flag.leverage <- diag_df$.hat > 4/nrow(diag_df)
diag_df$id <- 1:nrow(diag_df)


# Index plot for leverage
ggplot(data = diag_df, aes(x = reorder(id, FUN = identity, .hat), y = .hat, color = flag.leverage)) +
  geom_point() +
  labs(x = "Row index", y = "Leverage") +
  coord_flip()
```


# Arranging leverage

You can also reorder the data frame by the leverage values

```{r}
arrange(diag_df, desc(.hat)) %>%
  head(n = 10)
```

# Plotting Cook's distance

```{r}
# Flag Cook's D > 4/(n-2)
diag_df$flag.cooksd <- diag_df$.cooksd > 4/(nrow(diag_df) - 2)

# Index plot for Cook's D
ggplot(data = diag_df, aes(x = reorder(id, FUN = identity, .cooksd), y = .cooksd, color = flag.cooksd)) +
  geom_point() +
  labs(x = "Row index", y = "Cook's distance") +
  coord_flip()
```

# Arranging Cook's distance

```{r}
arrange(diag_df, desc(.cooksd)) %>%
  head(n = 10)
```