---
title: 'MLR: Model Selection in R'
author: "Math 430, Winter 2017"
output:
  ioslides_presentation
---

```{r, include=FALSE}
library(ggplot2)
library(car)
library(alr4)
library(plyr)
```

# Highway accident data

## {.smaller}

Variable    | Description
--------| -------------------------
`adt`   | average daily traffic count (thousands)
`trks`  | truck volume as a percent of the total volume
`lane`  | total number of lanes of traffic
`acpt`  | number of access points per mile
`sigs`  | number of signalized interchanges per mile
`itg`   | number of freeway-type interchanges per mile
`slim`  | speed limit
`len`   | length of the Highway segment (miles)
`lwid`  | lane width (feet)
`shld`  | width in feet of outer shoulder on the roadway
`htype` | type of roadway/funding source
`rate`  | accident rate per million vehicle miles


## {.centered}

```{r, fig.width=6.5, fig.height=6.5, echo=FALSE}
pairs(rate ~ ., data = Highway)
```

##

```{r echo = FALSE, fig.width = 8, fig.height = 6}
full_mod <- lm(rate ~ ., data = Highway)
residualPlots(full_mod, type = "rstandard", layout = c(3, 4), tests = FALSE)
```

## 

```{r echo = FALSE, fig.width = 8, fig.height = 6}
avPlots(full_mod, terms = ~ adt + trks + lane + acpt + sigs + itg + slim + len + lwid + shld, layout = c(3, 4))
```


##

```{r echo = FALSE, fig.width = 8, fig.height = 6}
mmps(full_mod, ~ adt + trks + lane + acpt + sigs + itg + slim + len + lwid + shld, layout = c(3, 4))
```


## {.smaller}

```{r echo=FALSE}
inverseResponsePlot(full_mod)
```


## {.smaller}

```{r include=FALSE}
# Investigate tforms using Box-Cox
summary(powerTransform(cbind(adt, trks, lane, acpt, slim, len, lwid, shld) ~ 1, data = Highway))
```


```{r tidy=TRUE}
Highway <- mutate(Highway, sigs1 = (sigs * len + 1) / len)
full_mod_tform <- lm(log(rate) ~ log(adt) + log(trks) + I(1/lane) + log(acpt) + itg + slim + log(len) + poly(lwid, 2) + poly(shld, 2) + htype + log(sigs1), data = Highway)
```

## 

```{r echo = FALSE, fig.width = 8, fig.height = 6}
residualPlots(full_mod_tform, type = "rstandard", layout = c(3, 4), tests = FALSE)
```

## 

```{r echo = FALSE, fig.width = 8, fig.height = 6}
avPlots(full_mod_tform, terms = ~ . - htype, layout = c(3, 4))
```


##

```{r echo = FALSE, fig.width = 8, fig.height = 6, message=FALSE, warning=FALSE}
mmps(full_mod_tform, ~ . - htype, layout = c(3, 4))
```



# The `step` command

## Backward elimination {.smaller}


```{r, include=FALSE}
belim <- step(full_mod_tform, scope = list(lower = ~ 1), direction = "backward", trace = 0)
```


```{r, eval=FALSE}
belim <- step(full_mod_tform, scope = list(lower = ~ 1), direction = "backward")
```

```{r}
broom::tidy(belim)
```



## Forward selection {.smaller}

```{r, include=FALSE}
null_mod <- lm(log(rate) ~ 1, data = Highway)
f <- formula(full_mod)
fselect <- step(null_mod, scope = list(lower = ~ 1, 
upper = ~ log(adt) + log(trks) + lane + acpt + itg + slim + log(len) + 
    lwid + shld + htype + sigs1),
direction = "forward", trace = 0)
```

```{r, eval=FALSE}
null_mod <- lm(slog(rate) ~ 1, data = Highway)
fselect <- step(null_mod, scope = list(lower = ~ 1, 
upper = ~ log(adt) + log(trks) + lane + acpt + itg + slim + log(len) + 
    lwid + shld + htype + sigs1),
direction = "forward")
```

```{r}
broom::tidy(fselect)
```


## Stepwise selection {.smaller}

```{r, tidy=FALSE, include=FALSE}
step_hwy <- step(null_mod, scope = list(lower = ~ 1, 
upper = ~ log(adt) + log(trks) + lane + acpt + itg + slim + log(len) + 
    lwid + shld + htype + sigs1),
direction = "both", trace=0)
```

```{r, tidy=FALSE, eval=FALSE}
step_hwy <- step(null_mod, scope = list(lower = ~ 1, 
upper =  ~ log(adt) + log(trks) + lane + acpt + itg + slim + log(len) + 
    lwid + shld + htype + sigs1),
direction = "both")
```

```{r}
broom::tidy(step_hwy)
```


## Using BIC rather than AIC {.smaller}

```{r, include=FALSE}
belim_bic <- step(full_mod_tform, scope = list(lower = ~ 1), direction = "backward", 
                   k = log(nrow(Highway)), trace=0)
```


```{r, eval=FALSE}
belim_bic <- step(full_mod_tform, scope = list(lower = ~ 1), direction = "backward", 
                   k = log(nrow(Highway)))
```

```{r}
broom::tidy(belim_bic)
```



# The `regsubsets` command

## All subsets in R {.smaller}

```{r tidy=TRUE}
library(leaps)
regfit_full <- regsubsets(log(rate) ~ log(adt) + log(trks) + I(1/lane) + log(acpt) + itg + slim + log(len) + poly(lwid, 2) + poly(shld, 2) + htype + log(sigs1), data = Highway, 
                          method = "exhaustive", nvmax  = 11,
                          nbest = 1)
reg_summary <- summary(regfit_full)
```

## Investigating the results

```{r, fig.height = 5, fig.width = 5}
plot(regfit_full, scale = "adjr2")
```

## Investigating the results

```{r, fig.height = 5, fig.width = 5}
plot(regfit_full, scale = "bic")
```

## Another plot option

```{r, include=FALSE}

best.vars <- function(scale, value){
  if(unique(scale) %in% c("r2", "adjr2")) value == max(value)
  else value == min(value)
}

allsub.df <- data.frame(vars = rep(1:11, times = 3), scale = rep(c("r2", "adjr2", "bic"), each = 11), value = c(reg_summary$rsq, reg_summary$adjr2, reg_summary$bic))
allsub.df$scale <- factor(allsub.df$scale, c("r2", "adjr2", "bic"))
allsub.df <- ddply(allsub.df, .(scale), mutate, optim = best.vars(scale, value))
```


```{r, echo=FALSE, fig.width = 6.5, fig.height = 5.5}
qplot(x =vars, y = value, data = allsub.df, geom = c("point", "line")) +
  facet_wrap(~ scale, ncol = 2, scales = "free") +
  xlab("No. Of Variables") +
  ylab("Value") + 
  scale_x_continuous(breaks=1:11) + 
  geom_point(data = subset(allsub.df, optim == TRUE), colour = I("orange"))
```

