---
title: "Bootstrapping Regression Models"
author: "Math 430, Winter 2017"
output: ioslides_presentation
---

# Motivation

```{r include=FALSE}
library(car)
library(ggplot2)
data(Transact)
```

## Transactions data

Data on transaction times in branch offices of a large Australian bank.

Variable | Description
-------- | -----------
`time`   | total transaction time (minutes)
`t1`     | \# type 1 transactions
`t2`     | \# type 2 transactions

## 

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align='center'}
pairs(time ~ ., data = Transact)
```

##

```{r fig.width = 8, fig.height = 3}
transact_mod <- lm(time ~ t1 + t2, data = Transact)
residualPlots(transact_mod, type = "rstandard", 
              layout = c(1, 3), tests = FALSE)
```

## 

```{r fig.width = 7, fig.height = 3.5}
avPlots(transact_mod,layout = c(1,2))
```

## 

```{r fig.width = 8, fig.height = 3}
mmps(transact_mod, layout = c(1,3))
```

## {.smaller}

```{r}
ncvTest(transact_mod)
ncvTest(transact_mod, ~ t1)
ncvTest(transact_mod, ~ t2)
```






</div>

# The Bootstrap

## Why bootstrap?

- We need an alternative method when the assumptions are suspect, or where standard methods are not readily available

- Can be used to
    + compute standard errors
    + compute confidence intervals
    + conduct tests
    
## Case resampling bootstrap

1. Number cases in data set from 1 to $n$. 

2. Take a random sample <u>with replacement</u> of size $n$ from these numbers.

3. Create a new data set by pulling the rows (cases) from the original data set that were selected in the random sample.
    
4. Fit the regression model to this new data set and save the values of the estimated coefficients, or other summary statistics.

5. Repeat steps 2-4 *R* times.

We can build confidence intervals from this list of *R* sets of coefficients.

## Case resampling bootstrap in R {.smaller}

```{r, message=FALSE}
library(car) # For Boot function
transact_boot <- Boot(transact_mod, R = 1999, method = "case")
summary(transact_boot)
```

```{r}
# Compare to the original fit
broom::tidy(transact_mod)
```


## Case resampling bootstrap in R

```{r, fig.width = 8.5, fig.height = 3.5}
hist(transact_boot, col = "gray", layout = c(1,3))
```


## Bootstrap confidence intervals

**Method 1:** Normal

$$\text{Estimate} \pm z^*_{\alpha/2} SE$$

where we use the bias corrected estimate: $statistic - \widehat{bias}$

```{r}
confint(transact_boot, level = .95, type = "norm")
```


## Bootstrap confidence intervals

**Method 2:** CIs based on percentiles of the bootstrap distribution

$$T^*_{\text{(lower)}} \quad \text{to} \quad T^*_{\text{(upper)}}$$

```{r}
confint(transact_boot, level = .95, type = "perc")
```

## Bootstrap confidence intervals

**Method 3:** Bias corrected and accelerated (BC$_a$) CIs

- More complicated, so don't worry about the formula
- Usually performs better than the percentile bootstrap CI

```{r}
confint(transact_boot, level = .95, type = "bca")
```



## Residual bootstrap

1. Fit the regression model to get $\widehat{\bf y} = \bf X \widehat{ \boldsymbol\beta}$ and compute the residuals, $e_i = y_i - \widehat{y}_i$

2. Obtain a random sample, <u>with replacement</u>, from the residuals to get a new sample (${\bf e^\prime} = (e^*_1, \ldots, e^*_n)$)

3. Create simulated $y$ values by $\bf y^* = \bf X \widehat{ \boldsymbol\beta} + e^*$

4. Fit the regression model to the simulated $y^*$ values and save the values of the estimated coefficients, or other summary statistics

5. Repeat steps 2-4 R times


We can build confidence intervals from this list of *R* sets of coefficients.

## Residual bootstrap in R {.smaller}

Note: This is only for illustrating how to use R to run the residual bootstrap, as there is evidence of nonconstant variance.

```{r}
transact_boot2 <- Boot(transact_mod, R = 1999, method = "residual")
summary(transact_boot2) # print summary
confint(transact_boot2, level = .95, type = "bca") # calc. CIs
```

## Case vs. Residual

- The residual bootstrap assumes: (1) linearity, and (2) constant variance.

- If this is true, then the residual bootstrap can be more accurate than the case bootstrap.