---
title: "Basics of Matrix Algebra in R"
author: "Math 430, Winter 2017"
output:
  pdf_document: default
  html_document: default
---

## Vectors


### Defining

```{r}
 a <- c(1, 3, 2)
```

Note: `a` will be printed as a row vector, but R defines it as a column vector. This can be seen through transposition

### Transposing

```{r}
t(a)
```

```{r}
t(t(a))
```

### Scalar multiplication

```{r}
3.14 * a
```


### Vector addition

```{r}
b <- c(2, 8, 9)
a + b
```



## Matrices

### Defining

```{r}
A <- matrix(c(1, 3, 2, 2, 8, 9), ncol = 3)
A
```

```{r}
 A2 <- matrix(c(1, 3, 2, 2, 8, 9), ncol = 3, byrow = TRUE)
A2
```


### Transposing

```{r}
t(A)
```



### Scalar multiplication

```{r}
3.14 * A
```


### Addition

```{r}
B <- matrix(c(5, 8, 3, 4, 2, 7), ncol = 3, byrow = TRUE)
A + B
```


### Multiplication

#### Matrix multiplied by a vector

A tempting, but incorrect, approach:

```{r}
X <- matrix(c(1, 1, 1, 2, 4, 6), ncol = 2, byrow = FALSE)
b <- c(5, 8)
X * b
```

Instead, use the `%*%` operator

```{r}
X %*% b
```

#### Matrix multiplied by a matrix

```{r}
t(X) %*% X
```


## Inverting a matrix

```{r}
Y <- c(1.5, 2.3, 3.7)
XtX <- t(X) %*% X
XtXinv <- solve(XtX)
XtXinv %*% t(X) %*% Y
```


## Special matrices

### The identity

```{r}
diag(3)
```

