---
title: "Adding Categorical Predictors"
author: "Math 430, Winter 2017"
output:
  ioslides_presentation
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggthemes)
library(ggExtra)
library(effects)
library(broom)
```


## The data

`UN11.csv` contains national health, welfare, and education statistics for 210 places, mostly UN members, but also other areas like Hong Kong that are not independent countries.

Variable |	Description
-----        | -------------------
`region`     |	region of the world
`group`      |	`oecd`, `africa`, or `other`
`fertility`  |	number of children per woman
`ppgdp`      |	per capita gross domestic product (US$)
`lifeExpF`   |	female life expectancy (years)
`pctUrban`   |	% urban



## The data {.smaller}

```{r}
UN11 <- read.csv("https://github.com/math430-lu/data/raw/master/UN11.csv")
head(UN11)
```

## Problem overview

**Primary research question**

- How does the expected life span of women (`lifeExpF`) differ between the three groups of countries?


**Analysis**

- What is the response variable?
- What should we use for the predictor?


## Life expectancy by group

```{r fig.width=4, fig.height=4}
UN11$group <- factor(UN11$group, levels = c("oecd", "other", "africa"))
ggplot(data = UN11, aes(x = group, y = lifeExpF)) + 
  geom_boxplot()
```

## Categorical predictors

- We use **dummy variables** (i.e. indicator variables) to put categories in a mathematical formula

- General idea:

<br>
<br>

- Coding `group` into dummy variables

## Fitting the no-intercept model in R {.smaller}

Model formula:

```{r}
lexp_grp <- lm(lifeExpF ~ group - 1, data = UN11)
summary(lexp_grp)
```

## Interpreting the coefficients

## Interpreting the coefficients {.smaller}

```{r fig.height = 4, fig.width = 4}
plot(Effect("group", mod = lexp_grp))
```

## Fitting the model with an intercept in R {.smaller}

```{r}
lexp_grp <- lm(lifeExpF ~ group, data = UN11)
summary(lexp_grp)
```

## Interpreting the coefficients

## Comparing level means

## Comparing level means

```{r}
vcov(lexp_grp)
```


## Adding a continuous predictor

**Research question**

- Can we find a model that better explains expected life span by incorporating other predictors?

**Analysis**

- MLR model with `group` and other predictors

## Adding a continuous predictor {.smaller}

```{r fig.width=5.5, fig.height=3.5}
ggplot(data = UN11, aes(x = log(ppgdp), y = lifeExpF, color = group, shape = group)) +
  geom_point() +
  scale_color_colorblind()
```

How can we build this model?

## Parallel lines model

$Y = \beta_0 + \beta_1 I_{other} + \beta_2 I_{africa} + \beta_3 x + e$

```{r echo=FALSE, fig.width=5.5, fig.height=3.5}
lexp_mod2 <- lm(lifeExpF ~ group + log(ppgdp), data = UN11)
parallel_df <- augment(lexp_mod2)

ggplot(data = UN11, aes(x = log(ppgdp), y = lifeExpF, color = group, shape = group)) +
  geom_point() +
  geom_line(data = parallel_df, aes(x = log.ppgdp., y = .fitted)) +
  scale_color_colorblind()
```


## Fitting the model in R {.smaller}

```{r}
parallel_mod <- lm(lifeExpF ~ group + log(ppgdp), data = UN11)
summary(parallel_mod)
```

## Interpreting the coefficients {.smaller}

```{r fig.height = 4, fig.width=8}
plot(allEffects(parallel_mod, default.levels=50), ylim=c(60, 85),
     grid=TRUE, multiline=TRUE)
```


## Unrelated lines model

$Y = \beta_0 + \beta_1 I_{other} + \beta_2 I_{africa} + \beta_3 x + \beta_4 I_{other} x + \beta_5 I_{africa} x + e$

```{r echo = FALSE, fig.width=5.5, fig.height=3.5}
ggplot(data = UN11, aes(x = log(ppgdp), y = lifeExpF, color = group, shape = group)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
  scale_color_colorblind()
```


## Fitting the model in R {.smaller}

```{r}
unrelated_mod <- lm(lifeExpF ~ group * log(ppgdp), data = UN11)
summary(unrelated_mod)
```

## Interpreting the coefficients {.smaller}

```{r fig.height = 5, fig.width=5}
plot(Effect(c("group", "ppgdp"), unrelated_mod, default.levels=100),
     rug=FALSE, grid=TRUE, multiline=TRUE)
```


## Comparing models

How can we determine which model we should prefer?

## Partial F-tests

## Partial F-tests in R

```{r}
anova(parallel_mod, unrelated_mod)
```

