---
title: "Inference for Regression Coefficients"
author: "Math 430, Winter 2017"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(ggplot2)
```

# Climate data

- Measurements on CO$_2$ in the atmosphere and global temperature anomaly (deviation from the mean temperature from 1961 to 1990)

```{r}
climate <- read.csv("https://github.com/math430-lu/data/raw/master/climate.csv")
head(climate)
```


- Goals 
    + understand the relationship between CO$_2$ and global temperatures
    
    + make predictions
    
---

# Climate data

```{r fig.height=5, fig.width=5}
ggplot(data = climate, mapping = aes(x = co2, y = globaltemp)) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE)
```


---

# Fitting the SLR model

```{r size="scriptsize"}
climate.lm <- lm(globaltemp ~ co2, data = climate)
summary(climate.lm)
```

---
class: middle, center

# Inference for the slope

---

# Statistical inference

**Goal:** use statistics calculated from data to makes inferences about the nature of parameters.

- Statistics: $\widehat{\beta}_0$, $\widehat{\beta}_1$

- Parameters: $\beta_0$, $\beta_1$


**Tools:**

- Confidence intervals
 
- Hypothesis tests


---

# Overview of statistical inference

---

# Confidence intervals

**Idea**: A confidence interval expresses the amount of uncertainly we have in our estimate of a particular parameter. 

To find such a range of plausible values for the parameter of interest, $\theta$, so that we know the long-run properties of the intervals.

$$P(\widehat{\theta}_L < \theta < \widehat{\theta}_U) = 1 - \alpha$$

- The endpoints are random variables **before** observing the data

- $\theta$ is fixed but unknown

---

# Confidence intervals

A general $1 - \alpha$ confidence interval takes the form

$$\widehat{\theta} \pm t^* \cdot SE(\widehat{\theta})$$

- $\alpha$: the confidence level

- $\widehat{\theta}$: a statistic (i.e. point estimate)

- $t^*$: the $1-\alpha/2$ quantile of a reference distribution

- $SE(\widehat{\theta})$: the standard error of $\widehat{\theta}$; i.e. the standard deviation of the sampling distribution

---

# Sampling distribution of the slope

Assume that the true model is

$$E(Y|X=x) = 12 + .7x, \quad e \sim \mathcal{N}(0, 4)$$
```{r, echo=FALSE,eval=TRUE, fig.width=5, fig.height=5}
n <- 60
beta_0 <- 12
beta_1 <- .7
sigma <- 2
x <- rnorm(n, mean = 20, sd = 3)

plot(20, 25, xlim = c(12, 28), ylim = c(17, 35), ylab = "y", xlab = "x", type = "n") # set up an empty plot
abline(a = beta_0, b = beta_1, col = "orange", lwd = 2) # add mean function
```

```{r eval = FALSE, echo = FALSE}
# generate data
f_mean <- beta_0 + beta_1 * x # mean function
f_data <- f_mean + rnorm(n, mean = 0, sd = sigma) # data generating function

points(x, f_data, pch = 16, col = "steelblue") # add generated data

# try to recover the true mean function
m1 <- lm(f_data ~ x)
m1
abline(m1)
```

---

# Sampling distribution of the slope

```{r, echo=FALSE,eval=TRUE, cache=TRUE}
it <- 5000
coef_mat <- matrix(rep(NA, it * 2), ncol = 2)
for(i in 1:it) {
  x <- rnorm(n, mean = 20, sd = 3)
  f_mean <- beta_0 + beta_1 * x
  f_data <- f_mean + rnorm(n, mean = 0, sd = sigma)
  coef_mat[i, ] <- lm(f_data ~ x)$coef
}
```

```{r, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7}
plot(20, 25, xlim = c(12, 28), ylim = c(17, 35), ylab = "y", xlab = "x", type = "n") # set up an empty plot

points(x, f_data, pch = 16, col = "steelblue") # add generated data
for(i in 1:it) {
  abline(coef_mat[i, 1], coef_mat[i, 2], col = rgb(0, 0, 0, 0.02))
}
abline(a = beta_0, b = beta_1, col = "orange", lwd = 2) # add mean function
```

---

# Sampling distribution of the slope

```{r, echo=FALSE, fig.height=7, fig.width=14}
beta_1s <- coef_mat[, 2]
hist(beta_1s, breaks = 20, 
     main = expression(paste("Sampling distribution of ", hat(beta)[1])),
     xlab = expression(hat(beta)[1]), prob = TRUE)
# mean(beta_1s)
# sd(beta_1s)
```

---

# Sampling distribution of the slope

```{r, echo=FALSE, fig.height=7, fig.width=14}
beta_1s <- coef_mat[, 2]
hist(beta_1s, breaks = 20, 
     main = expression(paste("Sampling distribution of ", hat(beta)[1])),
     xlab = expression(hat(beta)[1]), prob = TRUE)
abline(v = beta_1, col = "orange", lwd = 2)
text(beta_1+.02, .5, expression(beta[1]))
```

---

# Sampling distribution of the slope

```{r, echo=FALSE, fig.height=7, fig.width=14}
beta_1s <- coef_mat[, 2]
hist(beta_1s, breaks = 20, 
     main = expression(paste("Sampling distribution of ", hat(beta)[1])),
     xlab = expression(hat(beta)[1]), prob = TRUE)
abline(v = beta_1, col = "orange", lwd = 2)
text(beta_1+.02, .5, expression(beta[1]))
xl <- seq(min(beta_1s), max(beta_1s), length.out = 100)
lines(xl, dnorm(xl, mean = beta_1, sd = sqrt(sigma^2/sum((x-mean(x))^2))), 
      lwd = 2, col = "darkorange")
```

---

# Properties

1. $E(\widehat{\beta}_1 | X) = \beta_1$

1. $Var(\widehat{\beta}_1| X) = \dfrac{\sigma^2}{SXX}$

1. $\widehat{\beta}_1| X \sim \mathcal{N}\left(\beta_1,  \dfrac{\sigma^2}{SXX}\right)$

---

# Approximating the sampling distribution

We don't know $\sigma^2$, so we have to plug in our best guess at it's value, $S^2 = RSS/(n-2)$.

- The distribution is no longer normal due to the added uncertainty (heavier tails)

- Use the $t$ distribution with $n-2$ degrees of freedom (d.f.)

- Use R to find the quantiles

```{r eval=FALSE}
qt(1 - alpha/2, df = n-2)
```


---

# CI for the slope

$$\widehat{\beta}_1 \pm t^*_{\alpha/2, n-2} \dfrac{S}{\sqrt{SXX}}$$

---
class: middle

```{r}
summary(climate.lm)
```

---

# An easier way in R

```{r}
confint(climate.lm, level = 0.95)
```

---

# Interpreting CIs

We are 95% confident that the true slope between x and y lies between LB and UB.


For our climate example:

---

# Hypothesis testing framework

1. Formulate two competing hypotheses: $H_0$ and $H_A$.

1. Choose a test statistic that characterizes the information in the
sample relevant to $H_0$.

1. Determine the sampling distribution of the chosen statistic when $H_0$
is true.

1. Compare the calculated test statistic to the sampling distribution to determine whether it is "extreme."

---

# Tests for the slope

**Competing Claims:** $H_0: \beta_1 = \beta_1^0$ vs. $H_a: \beta_1 \ne \beta_1^0$

(R assumes that $\beta_1^0 = 0$)

<br>
<br>

**Test statistic:** $T = \dfrac{\widehat{\beta}_1 - \beta_1^0}{SE(\widehat{\beta}_1)}$
<br>
<br>

**Reference distribution**: $T \sim t_{n-2}$when $H_0$ is true

<br>
<br>

---

# In R

```{r}
summary(climate.lm)
```

---

# p-values

---
class: middle, center

# Inference for the intercept

---

# Sampling distribution of the intercept

$\widehat{\beta}_0 | X \sim \mathcal{N} \left(\beta_0,\  \sigma^2 \left( \dfrac{1}{n} + \dfrac{\overline{x}^2}{SXX} \right) \right)$

Again, we can only estimate $\sigma^2$ using $S^2 = RSS/(n-2)$



---

# Inference for the intercept

**Test statistic:** 

$T = \dfrac{\widehat{\beta}_0 - \beta_0^0}{se(\widehat{\beta}_0)}$, where $se(\widehat{\beta}_0) = S \sqrt{\dfrac{1}{n} + \dfrac{\overline{x}^2}{SXX}}$
<br>
<br>

**$(1-\alpha)100$% CI:**

$$\widehat{\beta}_0 \pm t^*_{\alpha/2, n-2} \cdot se(\widehat{\beta}_0)$$

---

# Centered SLR Model

**Issue:** $\beta_0$ is usually not interpretable

**Solution:** Center the predictor variable
$$x_i^* = x_i - \overline{x}$$
and fit the model
$$E(Y_i | X ) = \beta_0^* + \beta_1^* x_i^*$$

**Advantages:**

- Intercept is now the average/predicted value of $y$ when $x_i = \overline{x}$

- Slope and residual standard deviation stay the same

---

# Centering a variable in R


```{r message=FALSE}
library(dplyr)
climate <- mutate(climate, co2.center = co2 - mean(co2))
head(climate)
```


---

# Inference for $\beta_0$ in Centered SLR Model

**Test statistic:** 

$t = \dfrac{\widehat{\beta}_0^* - \beta_0^*}{se(\widehat{\beta}_0^*)}$, where $se(\widehat{\beta}_0^*) = S / \sqrt{n}$

<br>
<br>

**$(1-\alpha)100$% CI:**

$$\widehat{\beta}_0^* \pm t^*_{\alpha/2,n-2} \cdot se(\widehat{\beta}_0^*)$$

---

# Centered SLR Model in R

```{r}
centered.lm <- lm(globaltemp ~ co2.center, data = climate)
summary(centered.lm)
```
