---
title: "Introduction to regression"
author: "Prof. Adam Loy"
date: "Math 430, Winter 2017"
output: 
  ioslides_presentation:
    fig_height: 4
    fig_width: 4
    out_width: 400px
    fig_retina: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is a statistical model?

- A model is a representation for a particular purpose (Daniel Kaplan)
- A simplification of reality (Cannon et al., 2010)
- A statistical model embodies a set of assumptions concerning the generation of the observed data, and similar data from a larger population (Wikipedia)
- Remember that all models are wrong; the practical question is how wrong do they have to be to not be useful (George Box)

## What is a statistical model?

### Notation
- $Y_i$: **response variable** measured on the $i^{\text{th}}$ *observational unit*, $i=1,\ldots,n$
- $x_{i1}, x_{i2}, \ldots, x_{ik}$: **predictor/explanatory/covariate variables** measured on the $i^{\text{th}}$ *observational unit*

### Statistical model

$$Y_i = f(x_{i1}, x_{i2}, \ldots, x_{ik}) + \varepsilon$$

where

- $f(\cdot)$: a mathematical function
- $\varepsilon$: a random error term

## Why do we model?

1. Prediction

2. Understanding relationships

3. Assessing differences


## Challenge: Data collection and variable selection

- What data do we need to achieve our goals?
- Can we use data to actually answer our questions?
- What types of variables do we have?
- What variables are actually important?



## Challenge: Model specification

- What is the "right" model to use? 
- Does this choice help us achieve our goals?
- What assumptions are embedded in our choice?
- Are these assumptions valid?


## Challenge: Model specification {.columns-2}

<br>
```{r, echo=FALSE}
library(Stat2Data)
library(ggplot2)
data(TextPrices)

p <- qplot(x = Pages, y = Price, data = TextPrices) + 
  ylab("Textbook price") + 
  xlab("No. of pages")
print(p)
```

<br>
<br>
<br>
<br>
Need to choose $f(x)$


## Challenge: Model specification {.columns-2}

<br>
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(fpp)
library(forecast)
data(ausbeer)

ausbeer.df <- data.frame(quarter = as.numeric(time(ausbeer)), production = as.numeric(ausbeer))

ausbeer.arima <- auto.arima(ausbeer)
ausbeer.df$arima <- as.numeric(fitted(ausbeer.arima))

qplot(x = quarter, y = production, data = ausbeer.df) + 
  geom_line(aes(y = arima), colour = I("#377eb8"), alpha = I(0.8)) +
  geom_smooth(se = FALSE, colour = I("#e41a1c"), alpha = I(0.8)) + 
  ylab("Beer production") + 
  xlab("Year")
```

<br>
<br>
<br>
<br>
Need to choose $f(x)$


## Challenge: Model fitting {.columns-2}

```{r, echo=FALSE, message=FALSE}

library(MASS)
rmod <- rlm(Price ~ Pages, data = TextPrices)

p + geom_smooth(method = "lm", se = FALSE, colour = I("#377eb8"), linetype = 1) + 
  geom_abline(intercept = coef(rmod)[1], slope = coef(rmod)[2], colour = I("#e41a1c"), linetype = 2) + 
  geom_abline(intercept = -10, slope = .15, colour = I("#4daf4a"), linetype = 4)

```

<br>
<br>
<br>
<br>
How can we estimate the unknown parameters in our model?

$$f(x) = \beta_0 + \beta_1x$$

What should $\beta_0$ and $\beta_1$ be?


## Challenge: Dealing with randomness

If our model is 

$$Y_i = f(x_{i1}, x_{i2}, \ldots, x_{ik}) + \varepsilon$$

what distribution should $\varepsilon$ follow?


## What are we going to do in Math 430?

- Learn the statistical modeling thought process 

    1. Stating the problem in non-statistical terms
    2. Assessment of data quality
    3. Choose (specify) the model
    4. Fit the model
    5. Assess (validate) the model
    6. Use the model to answer the question(s)
    
- We will focus on regression models
    
## Logistics

### Textbook
- *A Modern Approach to Regression with R* by Simon J. Sheather, Springer, 2009
- One copy on 2-hour reserve at the Mudd

### Class sessions
- 2 days per week will focus on learning new methods
- Real examples emphasized

### Lab sessions
- 1 day each week will focus on making sure that you are comfortable with using R for statistical modeling


## Grading

### Homework (50%)
- Approximately weekly assignments
- Typically due Wednesdays by 4:30 p.m.
- No late work 

### Midterm (20%)
- To ensure that everyone is keeping up with the material

### Final project (30%)
- Carry out an entire analysis of a data set you choose
- More information later

## Overview {.columns-2}

### Topic
1. Simple linear regression
2. Multiple linear regression
3. Logistic regression

### Chapters
- 2--4
- 5--7
- 8
