---
title: "Introduction to R"
output:
  html_document: default
---

## Overview of R Markdown files

In this course we will often work with R markdown files in class. 

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd/Cntrl+Shift+Enter*. 

```{r}
1:10
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd/Cntrl+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).


## Your First R session

While you are learning the R language, remember that you are learning a new language; thus, we will start rather simply with small analysis tasks and build up to more complicated tasks. Also, you will not remember everything immediately---that's OK, it's a natural part of learning a language! 

### Installing and loading R packages

R does not enable all of its functionality when you open it. To enable additional functionality we need to load *R packages*. In this class we will often use the `tidyverse` R package to enable better tools for plotting, data manipulation, etc. Below is an example of installation and loading:

```{r message=FALSE, eval=FALSE}
# You only need to install a package once
install.packages("tidyverse")
```

```{r}
# You will need to load the package in each R markdown notebook
library(tidyverse)
```


### Loading data and assignment

Regression models are fit to data sets, so data will play a central role to our applied work in this course. There are multiple ways to load data sets (and we'll learn more about them later), but we often need to load a `.csv` (comma separated value) file.

Today, we'll look at a data set containing information on the weight (pounds) and fuel economy (mpg) of passenger vehicles

```{r}
# Load the data and assign it a name
mpg <- read.csv("https://github.com/math430-lu/data/raw/master/mpg.csv")
```


### Data frames

The `mpg` object is our first example of a data frame, which is essentially a list of vectors. We can get a first glimpse of our data set in a few ways:

```{r}
# Printing the first 6 rows
# Note that missing values are denoted by NA
head(mpg)
```

```{r}
# Printing the last 6 rows
tail(mpg)
```

```{r}
# Looking at the number of rows and columns
dim(mpg)
```

```{r}
# Looking at the structure
str(mpg)
```

```{r}
# Getting a glimpse of the structure
glimpse(mpg)
```


```{r}
# looking at the summary
summary(mpg)
```

### Syntax: Function application

We have already used functions in R---for example, `summary` is a function---so you have seen the syntax, but we need to make sure we understand what's happening.

The basic pattern is

```{r eval=FALSE}
object_name <- function_name(arguments)
```

Since we often have a data set of interest, many of our functions will take the form

```{r eval=FALSE}
object_name <- function_name(data, arguments)
```

### Exploring a bivariate relationship

Many factors go into determining what gas mileage a car will achieve. For now, we will focus on the weight of a car. Note that it's generally understood that heavier cars will get worse fuel economy, but it is not clear how much of an increase in weight will lead to a decrease in fuel economy.

Notation for regression models:

$y_i =$ fuel economy for $i^{\text{th}}$ car

$x_i =$ weight of $i^{\text{th}}$ car

$i = 1, \ldots, n$

$n =$ sample size

#### Scatterplots

The best place to start exploring the relationship between two quantitative variables is with a scatterplot.

```{r, fig.height=4, fig.width=4}
ggplot(data = mpg, mapping = aes(x = weight, y = mpg)) + 
  geom_point() +
  labs(x = "weight (lbs.)", y = "fuel economy (mpg)")
```

When examining a scatterplot we must keep the following features in mind:

* Form
* Direction
* Strength
* Outliers / unusual features

#### Correlation

We can also quantify the strength of the linear relationship between two quantitative variables using correlation.

**Formula:**
$$Corr(X, Y) = r = \dfrac{1}{n-1} \sum_{i = 1}^n \left( \dfrac{x_i - \overline{x}}{s_x} \right) \left( \dfrac{y_i - \overline{y}}{s_y} \right)$$

**Key properties:**

  1. $-1 \le r \le 1$
  2. scale invariant
  
```{r}
cor(mpg$weight, mpg$mpg)
```

**Caution:**

Correlation only measures the strength/direction of **linear relationships**


```{r}
x <- -5:5
y <- x^2
cor(x, y)
```

```{r, echo=FALSE, fig.height=4, fig.width=4}
ggplot(data = data.frame(x, y), mapping = aes(x, y)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```


Correlation is **sensitive to outliers**

```{r, echo=FALSE, message=FALSE, fig.height=4, fig.width=8}
library(gridExtra)
p1 <- qplot(x3, y3, data = anscombe) +
  xlab("x") + ylab("y") + scale_y_continuous(limits=c(5,13)) +
  annotate("text", label = "r = 0.816", x = 6, y = 12, size = 6)

p2 <- qplot(x3, y3, data = subset(anscombe, y3 < 9)) +
  xlab("x") + ylab("y") + scale_y_continuous(limits=c(5,13)) +
  annotate("text", label = "r = 0.999", x = 6, y = 12, size = 6)

grid.arrange(p1, p2, nrow = 1)
```

